---
title: "Analysing the Key Factors in predicting the 2020 Election"
author: "Victor ChanYoung Cho, Usman Sadiq"
date: "Nov 2nd, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Abstract

TODO The overall summmary of the report; Mainly include background, what we did for the analyses, results and discussion.

# Introduction

An overarching issue that rises an immense political controvery is the United States presidential election. In 2016, Donald Trump impressed the political scene by being elected as the Republican nominee for president. He also continuously won the white house votes, which eventually lead to a surprise of being elected as the president, while many observers of the political scene predicted the election of Hillary Clinton. (Jérôme et. al, 2020) Now, with the election being an ongoing event, Trump has already checked on the different parties in the previous years. (Kim & Kim, 2020) With the ongoing presidential debate as well, Trump has also been having his firm stance against Biden, for his another successful election. The US presidential election as we observe is mainly a competition of the current and former president versus the most popular candiate. We will be modelling a multilevel logistic regression with the UCLA nationscape dataset and preform a poststratification process with the American Community Surveys (ACS) to predict the election outcome and to target the most successful candidate for the 2020 US election.

# Data

We will be using the Democracy Fund + UCLA Nationscape ‘Full Data Set’ for modelling and the ACS 2018 5 year data for post-stratification in order to perform our analysis. From the ACS data, we selected the key factors which we believed would have the strongest effect on the prediction of the outcome. The variables are the region, state, sex, age, race, hispanic origin, birthplace, years in the united states, education, labor force status and total income.


```{r setup, include=FALSE}
library(tidyverse)

# Loading in the cleaned survey Data
survey_data <- read_csv("./outputs/survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("./outputs/census_data.csv")
```


# Model

Our goal for the study as mentioned above is to predict the outcome of the US election, particularly in the interest of Donald Trump's re-election. Using the Democracy Fund + UCLA Nationscape ‘Full Data Set’ from the Voters Study Group(@citeSurveyData), We will implement a binary logistic model that concerns whether citizens will vote from Trump or Bide, then analyse the proportions of Trump voters and Biden voters. 

To supplement the above, we will also be applying the multilevel regression post-stratification technique on the overall census data(@citeCensusData) obtained through IPUMS to yield the expected the overall outcome of the US election. The detailed procedure and outcome will be described below in the post-stratification section and the results.

## Model Specifics

We have implemented the linear regression model to model the proportion of the voters who will vote for Donald Trump. We decided to use the logistic regression model as we saw the election process as a binomial process. That is, whether people vote for Trump or Biden could be seen as one Bernoulli random variable, and multiple trials of these result in a binomial process. As mentioned in the data section, the variables we have taken into account to model the probability of voting for Donald Trump is region, state, sex, age, race, hispanic origin, birthplace, years in the united states, education, labor force status and total income. We present our logistic regression model by the following:


$$ log(p/1-p) = \beta_0+\beta_1  x_{employment} + \beta_2 x_{gender} + \beta_3 x_{race_ethnicity} + \beta_4 x_{household_income} + \beta_5 x_{education} + \beta_6 x_{age} + \beta_7 x_{hispanic} $$

Where $p$ represents the proportion of voters who will vote for Donald Trump. Similarly, $\beta_0$ represents the intercept of the model, and is the probability of voting for Donald Trump at age 0. $\beta_1$ represents the log odds that change for each unit increase in employment, $\beta_2$ represents the log odds based on gender, $\beta_3$  based on ethnicity, $\beta_4$ based on household income, $\beta_5$  based on education, $\beta_6$ based on age, and $\beta_7$, based on hispanic.


```{r, include=FALSE}

logreg <- glm(data=survey_data, vote_trump~as.factor(gender)+as.factor(race_ethnicity)+as.factor(household_income)+as.factor(education)+age, family = "binomial")

```

## Post-Stratification 

In order to estimate the proportion of voters who will vote for Donald Trump I need to perform a post-stratification analysis. Here I create cells based off different ages. Using the model described in the previous sub-section I will estimate the proportion of voters in each age bin. I will then weight each proportion estimate (within each bin) by the respective population size of that bin and sum those values and divide that by the entire population size. 

```{r, include=FALSE}

census_data

# Here I will perform the post-stratification calculation
census_data$logodds_estimate <-
  logreg %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))


census_data %>%
  mutate(alp_predict_prop = estimate*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))


```


# Results

Here you will include all results. This includes descriptive statistics, graphs, figures, tables, and model results. Please ensure that everything is well formatted and in a report style. You must also provide an explanation of the results in this section. 

```{r}
broom::tidy(logreg)
```


Please ensure that everything is well labelled. So if you have multiple histograms and plots, calling them Figure 1, 2, 3, etc. and referencing them as Figure 1, Figure 2, etc. in your report will be expected. The reader should not get lost in a sea of information. Make sure to have the results be clean, well formatted and digestible.

# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

## Weaknesses

Here we discuss weaknesses of the study, data, analysis, etc. You can also discuss areas for improvement.

## Next Steps

Here you discuss subsequent work to be done after this report. This can include next steps in terms of statistical analysis (perhaps there is a more efficient algorithm available, or perhaps there is a caveat in the data that would allow for some new technique). Future steps should also be specified in terms of the study setting (eg. including a follow-up survey on something, or a subsequent study that would complement the conclusions of your report).


# References

Tausanovitch, Chris and Lynn Vavreck. 2020. Democracy Fund + UCLA Nationscape, October 10-17, 2019 (version
  20200814). Retrieved from [https://www.voterstudygroup.org/downloads?key=9e0589c5-5ecc-4575-be2a-8285262f97fe.

Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose Pacas and Matthew Sobek. IPUMS USA: Version
  10.0 [dataset]. Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D010.V10.0

Jérôme, B., Jérôme, V., Mongrain, P., & Nadeau, R. (2020). State-Level Forecasts for the 2020 US Presidential 
  Election: Tough Victory Ahead for Biden. PS: Political Science & Politics, 1-4. doi:10.1017/S1049096520001377

Kim, A., &amp; Kim, P. (2019). Estimation of the 2020 US Presidential Election Competition and Election 
  Stratagies. 2019 IEEE 10th Annual Ubiquitous Computing, Electronics &amp; Mobile Communication Conference 
  (UEMCON). doi:10.1109/uemcon47517.2019.8992973
  