---
title: "Predicting the 2020 Election and Analysing the Key Factors"
author: "Victor ChanYoung Cho, Usman Sadiq"
date: "Nov 2nd, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Abstract

In the midst of the 2020 United States Election issue, we took this opportunity to investigate the leading factors and predict its outcome using the Daemocracy Fund + UCLA Nationscape ‘Full Data Set’ data (Survey data) and American Community Survey 2018 5 year data (Census Data). After filtering out the approrpriate variables and categorising the varying responses, we modelled a logistic regression equation with the survey data, and analysed the significance of the variables. After that, with the logisitic equation model that we have built, we used the census data to conduct a post-stratification process to finally predict the probability that Donald Trump would win the election. Results showed that the variables we have selected (age, gender, household income to name a few) were all significant key factors that may impact the 2020 US election, and we consequently yielded a result where Donald Trump is likely to not win the US Election. Some limitations exist with our analyses however, mainly such that our varying categorical responses were filtered and simplified based on their relative similarity. We address this bias may have impacted our results, and we suggest selecting numerical variables for the regression model for a better result for future work.



# Introduction

An overarching issue that rises an immense political controvery is the United States presidential election. In 2016, Donald Trump impressed the political scene by being elected as the Republican nominee for president. He also continuously won the white house votes, which eventually lead to a surprise of being elected as the president, while many observers of the political scene predicted the election of Hillary Clinton. (Jérôme et. al, 2020) Now, with the election being an ongoing event, Trump has already checked on the different parties in the previous years. (Kim & Kim, 2020) With the ongoing presidential debate as well, Trump has also been having his firm stance against Biden, for his another successful election. The US presidential election as we observe is mainly a competition of the current and former president versus the most popular candiate. We will be modelling a multilevel logistic regression with the UCLA nationscape dataset and preform a poststratification process with the American Community Surveys (ACS) to predict the election outcome and to target the most successful candidate for the 2020 US election.

# Data

We will be using the Democracy Fund + UCLA Nationscape ‘Full Data Set’ for modelling and the ACS 2018 5 year data for post-stratification in order to perform our analysis. From the ACS data, we selected the key factors which we believed would have the strongest effect on the prediction of the outcome. The variables are the region, state, sex, age, race, hispanic origin, birthplace, years in the united states, education, labor force status and total income.


```{r setup, include=FALSE}
library(tidyverse)

# Loading in the cleaned survey Data
survey_data <- read_csv("./outputs/survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("./outputs/census_data.csv")
```


# Model

Our goal for the study as mentioned above is to predict the outcome of the US election, particularly in the interest of Donald Trump's re-election. Using the Democracy Fund + UCLA Nationscape ‘Full Data Set’ from the Voters Study Group, We will implement a binary logistic model that concerns whether citizens will vote from Trump or Bide, then analyse the proportions of Trump voters and Biden voters. 

To supplement the above, we will also be applying the multilevel regression post-stratification technique on the overall census data obtained through IPUMS to yield the expected the overall outcome of the US election. The detailed procedure and outcome with our built model will be described below in the post-stratification section and the results section.

## Model Specifics

We have implemented logistic regression to model the proportion of the voters who will vote for Donald Trump. We decided to use this model as we saw the election process as a binomial process. That is, whether people vote for Trump or Biden could be seen as one Bernoulli random variable, and multiple trials of these result in a binomial process. As mentioned in the data section, the variables we have taken into account to model the probability of voting for Donald Trump is gender, race_ethnicity, household_income, education, and age. We present our logistic regression model by the following:


$$ log(p/1-p) = \beta_0 +\beta_1  x_{gender} + \beta_2 x_{race_enthnicity} + \beta_3 x_{household_income} + \beta_4 x_{education} + \beta_5 x_{age} $$

Where $p$ represents the proportion of voters who will vote for Donald Trump. $\beta_0$ represents the intercept of the model. The components consist of $\beta_1$ representing the log odds based on gender, $\beta_2$ based on race ethnicity, $\beta_3$  based on household income, $\beta_4$ based on age, and $\beta_5$ based on education.

```{r, include=FALSE}

logreg <- glm(data=survey_data, vote_trump~as.factor(gender)+as.factor(race_ethnicity)+as.factor(household_income)+as.factor(education)+age, family = "binomial")

broom::tidy(logreg)



```

## Post-Stratification 

In order to estimate the proportion of voters who will vote for Donald Trump I need to perform a post-stratification analysis. Here I create cells based off different ages. Using the model described in the previous sub-section I will estimate the proportion of voters in each age bin. I will then weight each proportion estimate (within each bin) by the respective population size of that bin and sum those values and divide that by the entire population size. 

```{r, include=FALSE}

census_data

# Here I will perform the post-stratification calculation
census_data$logodds_estimate <-
  logreg %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))


census_data %>%
  mutate(alp_predict_prop = estimate*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))


```


# Results

Here you will include all results. This includes descriptive statistics, graphs, figures, tables, and model results. Please ensure that everything is well formatted and in a report style. You must also provide an explanation of the results in this section. 

```{r}
broom::tidy(logreg)
```


Please ensure that everything is well labelled. So if you have multiple histograms and plots, calling them Figure 1, 2, 3, etc. and referencing them as Figure 1, Figure 2, etc. in your report will be expected. The reader should not get lost in a sea of information. Make sure to have the results be clean, well formatted and digestible.

# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

## Weaknesses

While our analyses showed plausible results, it is important to understand that the prediction must be taken as a rough estimate. Preparing the data was not easy as there were so many categorical variables. Due to the above, we had to immensely simplify the varying responses of the categorical variables into constrained groupings. For instance, relatively similar responses were grouped into their own categories and data given with a range of values were mutated so the census data and the survey data match up. We address that this could have potentially caused some inaccuracy, as our data used such categorically constrained variables to simplify the categorical responses.

## Next Steps

We feel that is necessary to compare our analyses results to the actual outcome of the election. This way, not only we maybe able to analyse our possible caveats that may have been missed. We could possibly apply this to furthur analyses after the election, and analyse what may have been the biggest factor behind the results.

We assertthat it would be a lot better if the variables to be used for the dataset would be selected differently. We feel that due to varying categorical responses, it was hard for us to find a common ground in some groupings and we heavily felt that some categorical groupings may be subjective. If our model contained more non-categorical variables, we feel that our results will not be biased towards the categories we have made based on our relative judgments on which response should grouped with while simplying our data.

# References

Tausanovitch, Chris and Lynn Vavreck. 2020. Democracy Fund + UCLA Nationscape, October 10-17, 2019 (version
  20200814). Retrieved from [https://www.voterstudygroup.org/downloads?key=9e0589c5-5ecc-4575-be2a-8285262f97fe.

Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose Pacas and Matthew Sobek. IPUMS USA: Version
  10.0 [dataset]. Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D010.V10.0

Jérôme, B., Jérôme, V., Mongrain, P., & Nadeau, R. (2020). State-Level Forecasts for the 2020 US Presidential 
  Election: Tough Victory Ahead for Biden. PS: Political Science & Politics, 1-4. doi:10.1017/S1049096520001377

Kim, A., &amp; Kim, P. (2019). Estimation of the 2020 US Presidential Election Competition and Election 
  Stratagies. 2019 IEEE 10th Annual Ubiquitous Computing, Electronics &amp; Mobile Communication Conference 
  (UEMCON). doi:10.1109/uemcon47517.2019.8992973
  